Manuscript_Github_Pchabaudi_calculateTPM

### 2021-03-22 Manuscript_pirAnalysis_Github/Pchabaudi24h/Manuscript_Github_Pchabaudi_calculateTPM.Rmd
Pchabaudi TPM file

---
title: "Manuscript_Github_Pchabaudi_pir"
author: "Timothy Little"
date: "22/03/2021"
output: html_document
---

@@

```{r colours_clades_phase}
SorL_col <- setNames(
  c(    '#F0E442',
        '#56B4E9', #L
        '#E69F00', #S
        '#999999'),
  sort(
    unique(cir_info$SorL)
  )
)

SorL_col_num <- SorL_col
names(SorL_col_num) <- sapply(names(SorL_col), 
                              function(clade_name) {
                                paste0(
                                  clade_name, 
                                  ' (n = ',
                                  sum(cir_info$SorL == clade_name),
                                  ')'
                                )
                              })

clade_col <- setNames(
  c('#F0E442',
    '#56B4E9', #L1
    '#009E73',
    '#0072B2',
    '#E69F00', #S1
    '#D55E00',
    '#CC79A7',
    '#999999'),
  sort(
    unique(cir_info$subfam)
  )
)

clade_col_num <- clade_col
names(clade_col_num) <- sapply(names(clade_col), 
                               function(clade_name) {
                                 paste0(
                                   clade_name, 
                                   ' (n = ',
                                   sum(cir_info$subfam == clade_name),
                                   ')'
                                 )
                               })

@@

## Figure 3c - Highest expressed all _pir_ heatmap

```{r highest_expressed_stage_heatmap_fig3b}

num_SorL <- cir_info %>% 
  group_by(SorL) %>% 
  summarise(num = n_distinct(Gene.ID))

labels_SorL <- setNames(paste0(num_SorL$SorL, ' (n = ', num_SorL$num, ')'),
                        num_SorL$SorL)

# Get order of pir genes by highest expressed genes in each sample.

combined_data_pir_ht_list <- lapply(names(twentyfour_avgtpm), 
       function(list_name){

         combined_data_pir <- getElement(twentyfour_avgtpm, 
                                     list_name)$tpm %>% 
           filter(Geneid %in% cir_id) %>% 
           column_to_rownames(var = 'Geneid')

         #Ordering that a more understandable order

         combined_data_pir <- combined_data_pir[,c(ordering(colnames(combined_data_pir)[-1], 
                                                            order_vector)),
                                                drop = FALSE]

         combined_data_ordered <- tibble(Geneid = names(apply(combined_data_pir, 1, mean)),
                                         means = apply(combined_data_pir, 1, mean),
                                         subfam = cir_info$SorL[match(names(apply(combined_data_pir, 1, mean)), 
                                                                            cir_info$Geneid)]) %>% 
           filter(means > 1) %>% 
           arrange(subfam, 
                   desc(means)) %>% 
           .$Geneid

         #Order the heatmap data by the order defined above (SorL, then mean expression).
         combined_data_pir_ht <- combined_data_pir[
           order(
             match(rownames(combined_data_pir),
                   combined_data_ordered),
             na.last = NA #Remove NAs
           ),
         ]

         #Change the colnames to the times
         colnames(combined_data_pir_ht) <- str_extract(colnames(combined_data_pir_ht),
                                                                           pattern = '[[:digit:]]{2}h')

         #Order by time point
         combined_data_pir_ht <- combined_data_pir_ht[,order(colnames(combined_data_pir_ht)),
                                                      drop = FALSE] 
         #Add a column for the SorL
         combined_data_pir_ht$SorL <- cir_info$SorL[match(names(apply(combined_data_pir_ht, 1, mean)), 
                                                                cir_info$Geneid)]

         return(combined_data_pir_ht)
       })
names(combined_data_pir_ht_list) <- names(twentyfour_avgtpm)

maxcol <- ceiling(
  max(
    unlist(
      lapply(combined_data_pir_ht_list,
             function(list_element){
               max(
                 log2(1+list_element[!colnames(list_element) %in% 'SorL'])
               )
             }
      )
    )
  )
)

lapply(names(combined_data_pir_ht_list),
       function(list_name_to_order){

         combined_data_pir_ht <- getElement(combined_data_pir_ht_list,
                                            list_name_to_order)

         pir_highest_tpm_order <- rownames(combined_data_pir_ht)

         #Sub-family annotation
         clade_info <- combined_data_pir_ht$SorL
         annotation_col <- SorL_col

         #Subfamily annotation

         clade_ha <- rowAnnotation(
           'sub-family' = anno_simple(
             clade_info,
             col = annotation_col,
           ),
           annotation_name_gp = gpar(fontsize = 10)
         )

         anc_ha <- rowAnnotation(foo = anno_mark(at =  na.omit(match(clade_info, 'ancestral')) , 
                                                 labels = 'ancestral',
                                                 labels_gp = gpar(fontsize = 7),
                                                 link_width = unit(3, "mm")))

         #lapply across the datasets now, for each order
         ht_list <- lapply(names(combined_data_pir_ht_list),
                           function(list_name){

                             combined_data_pir_ht <- getElement(twentyfour_avgtpm, 
                                                                list_name)$tpm %>% 
                               column_to_rownames(var = 'Geneid')

                             #Get the times instead of the stage names.
                             colnames(combined_data_pir_ht) <- str_extract(colnames(combined_data_pir_ht),
                                                                           pattern = '[[:digit:]]{2}h')

                             combined_data_pir_ht <- combined_data_pir_ht[,order(colnames(combined_data_pir_ht)),
                                                                          drop = FALSE]

                             Heatmap(
                               log2(1+combined_data_pir_ht)[pir_highest_tpm_order,],
                               cluster_rows = FALSE,
                               cluster_columns = FALSE,
                               show_row_names = FALSE,
                               name = 'log(TPM+1)', 
                               col = colorRamp2(c(0,
                                                  maxcol/4,
                                                  maxcol/2,
                                                  3*maxcol/4,
                                                  maxcol),
                                                viridis(5)),
                               na_col = 'white', 
                               border = TRUE,
                               border_gp = gpar(lwd = 0.1),
                               column_names_gp = gpar(fontsize = 10),
                               row_dend_width = unit(3, "cm"),
                               width = unit(4.5, 'cm'),
                               height = unit(7,'cm'),
                               row_title = NULL,
                               left_annotation = if(
                                 list_name %in% dplyr::first(names(combined_data_pir_ht_list))
                               ){
                                 clade_ha
                               }else{
                                 NULL
                               },
                               right_annotation = if(
                                 list_name %in% dplyr::last(names(combined_data_pir_ht_list))
                               ){
                                 anc_ha
                               }else{
                                 NULL
                               },
                               # row_split =  cir_info$SorL[match(pir_highest_tpm_order, 
                               #                                        cir_info$Geneid)],
                               column_title = paste(list_name),
                               heatmap_legend_param = list(
                                 labels_gp = gpar(fontsize = 10),
                                 title_gp = gpar(fontsize = 10,
                                                 fontface = 'bold'),
                                 at = c(0,
                                        maxcol/4,
                                        maxcol/2,
                                        3*maxcol/4,
                                        maxcol)
                               )
                             )
                           }
         )
         ht_list <- Reduce(`+`, ht_list)
         clade_lgd <- Legend(
           title = 'sub-family',
           at = labels_SorL,
           legend_gp = gpar(fill = SorL_col)
         )

         return(
           draw(ht_list, 
                row_title_side = 'left',
                column_title_side = 'bottom',
                heatmap_legend_side = 'right',
                annotation_legend_side = 'right',
                annotation_legend_list = list(clade_lgd),
                ht_gap = unit(.35,'cm'),
                merge_legend = TRUE
           ))
       })

## 2D plot of PC1 v PC2 - Supp figure 4A

```{r pca_calculation, results = TRUE}
twentyfour_tpm <- twentyfour_ct_comb
rownames_geneid <- twentyfour_tpm$Geneid
twentyfour_tpm_forpca <- select(twentyfour_tpm, -Geneid)
rownames(twentyfour_tpm_forpca) <- rownames_geneid
twentyfour_tpm_pca <- prcomp(twentyfour_tpm_forpca, center = TRUE, scale. = TRUE)
#Need to look into what exactly scale and center are doing
```

```{r 2d_pca, results = TRUE}
twentyfour_tpm_pca_var <- get_pca_var(twentyfour_tpm_pca)
twentyfour_tpm_pca_coords <- data.frame(twentyfour_tpm_pca_var$coord)

twentyfour_tpm_pca_coords <- data.frame(sample = rownames(twentyfour_tpm_pca_var$coord),
                                        PC1 = twentyfour_tpm_pca_coords$Dim.1,
                                        PC3 = twentyfour_tpm_pca_coords$Dim.3,
                                        PC2 = twentyfour_tpm_pca_coords$Dim.2) %>% 
  dplyr::mutate(stage = gsub(sample, pattern = '_[[:digit:]]$', replacement = '')) %>% 
  dplyr::mutate(time = str_split_fixed(stage, pattern = '_', n = 2)[,2]) %>% 
  dplyr::mutate(transmission = str_split_fixed(sample, pattern = '_', n = 3)[,1])

colornames <- viridis(length(unique(twentyfour_tpm_pca_coords$time)))
colornames <- setNames(colornames, 
                       sort(unique(twentyfour_tpm_pca_coords$time)))

pca_importance <- as.data.frame(summary(twentyfour_tpm_pca)$importance)
pc1_importance <- round(pca_importance$PC1[2]*100, digits = 1)
pc2_importance <- round(pca_importance$PC2[2]*100, digits = 1)
pc3_importance <- round(pca_importance$PC3[2]*100, digits = 1)

(
  plot <- ggplot(twentyfour_tpm_pca_coords,
                 aes(
                   x = PC1,
                   y = PC2,
                   col = time,
                   shape = transmission
                 )) +
    geom_point(size = 3,
               alpha = 0.8) +
    scale_color_manual(values = colornames) +
    scale_shape_manual(
      values = setNames(
        c(17,19),
        unique(twentyfour_tpm_pca_coords$transmission)
      )
    ) +
    xlab(paste0('PC1 ', round(pca_importance$PC1[2]*100, digits = 1),'%')) +
    ylab(paste0('PC2 ', round(pca_importance$PC2[2]*100, digits = 1),'%')) +
    guides(shape=guide_legend(ncol=2),
           col = guide_legend(ncol=2)) +
    theme_classic() +
    theme(legend.key.size = unit(4, 'mm'),
          legend.text = element_text(size = 7))
)

# Calculate and cluster with z-scores {.tabset}

z-score represents the number of standard deviations from the mean of the data. To calculate from TPM or counts for a given gene X:

1. Convert the count/RPKM values of each gene into log values.
2. Calculate the mean and standard deviation of X gene log values in each sample
3. For first sample: (gene X log value - mean of log values of samples)/ standard deviation of log values of all samples
4. Now. i have the z-score for gene x in first sample. Using the above protocol, I can convert all genes log values into z-score.

* Need to make sure that the z-score is calculated within each transmission method!

```{r zscores}
tpm_to_zscore = function(input){

  #If the input does not have a TPM element then assume that the input itself is the TPM object
  if(any(names(input) %in% 'tpm')){
    tpm_object <- input$tpm
  }else{
    tpm_object <- input
    warning('No \'tpm\' element in the supplied input so function will assume that the input is the tpm data')
  }

  #Want all of them so will just ignore Geneid
  input_logged <- tpm_object %>%
    mutate_at(vars(-matches('Geneid')), 
              log1p)

  gene_mean_scores = input_logged %>% 
    melt(id.vars = 'Geneid') %>% 
    group_by(Geneid) %>% 
    dplyr::summarise(mean = mean(value),
                     std_dev = sd(value))

  gene_vector = as.character(input_logged$Geneid)

  input_z = lapply(gene_vector, function(gene){
    mean_gene = filter(gene_mean_scores, 
                       Geneid == paste(gene)) %>% 
      select(mean)
    std_gene = filter(gene_mean_scores, 
                      Geneid == paste(gene)) %>% 
      select(std_dev)
    gene_scores = input_logged %>% 
      filter(Geneid == paste(gene)) %>% 
      select(-Geneid)
    if(std_gene != 0){
      gene_z = sapply(gene_scores, function(score){(score-mean_gene)/std_gene}) %>% data.frame
    }else{
      gene_z = (sapply(gene_scores, function(score){0})) %>% data.frame %>% t
    }
    colnames(gene_z) = colnames(gene_scores) #For some reason apply makes the column names weird
    gene_z = cbind(data.frame(Geneid = paste(gene)), gene_z)
  })

  #Use reduce and rbind to turn the list into one df
  input_z_df = purrr::reduce(input_z, 
                             rbind)
  return(list(zscores = input_z_df,
              tpm = tpm_object,
              experiment = input$experiment))
}

#Make and save z-scores
twentyfour_z <- lapply(twentyfour_avgtpm,
                       tpm_to_zscore)

###
```

# Phaseograms of all the genes in both transmission methods - supplementary figure 5.

```{r row_order_phaseograms}

#Getting the row order for each phaseogram for each experiment
# Want the data ordered by the highest zscore for each gene in each stage.
phaseogram_row_order <- lapply(twentyfour_z, 
                               function(list_elements){
                                 zscore_df <- list_elements$zscores
                                 zscore_ordered <- melt(zscore_df, 
                                                        id = 'Geneid',
                                                        variable.name = 'stage',
                                                        value.name = 'zscore') %>% 
                                   group_by(Geneid) %>% 
                                   top_n(zscore, n = 1) %>% 
                                   arrange(stage, 
                                           desc(zscore)) %>%
                                   distinct(Geneid, 
                                            .keep_all = TRUE) %>%  #Keep only the ones that are uniquely highest expressed in one sample
                                   filter(zscore > 0) #For some reason some remain with zero zscore
                                 # zscore_ordered <- zscore_ordered[order(
                                 #   match(
                                 #     zscore_ordered$stage, 
                                 #     order_vector
                                 #   )
                                 # ),]
                                 return(c(list_elements, 
                                          list(zscore_order = zscore_ordered)))
                               })

phaseogram_time_order <- lapply(names(phaseogram_row_order), 
                                function(list_element){
                                  transmission_list <- getElement(phaseogram_row_order,
                                                                  list_element)
                                  order_by_time <- transmission_list$zscore_order %>% 
                                    mutate(stage = times_stages_table$time[match(stage,
                                                                                 times_stages_table$r_stage_diff_count)]) %>% 
                                    arrange(stage)
                                  return(list(zscore_order = order_by_time))
                                }
)
names(phaseogram_time_order) <- names(phaseogram_row_order)

```{r function_draw_heatmap_zscore}

draw_heatmap_zscore <- function(zscore_list,
                                title = '',
                                gene_ordering_vector = NULL,
                                filter_genes_vector = NULL,
                                extra_title = NULL,
                                annotation_clade = NULL,
                                annotation_col,
                                threshold_tpm = NULL,
                                use_times = FALSE,
                                use_order_vector = TRUE,
                                tpm_mat = TRUE,
                                tpm_breaks = NULL,
                                ...){
  if(is.null(tpm_breaks) & tpm_mat){
    #Need the generate the breaks in the data for the heatmap legend if not provided.
    max_tpm <- lapply(zscore_list, 
                      function(list_element){
                        #list_element <- zscore_list[[1]]
                        heatmap_data <- as.data.frame(list_element$zscores)
                        title <- list_element$experiment

                        if(!'Geneid' %in% colnames(heatmap_data)){
                          stop('No Geneid column in the heatmap data.frame')
                        }

                        if(!is.null(filter_genes_vector)){
                          heatmap_data <- as.data.frame(
                            filter(heatmap_data, 
                                   Geneid %in% filter_genes_vector)
                          )
                        }

                        rownames(heatmap_data) <- as.character(heatmap_data$Geneid)

                        heatmap_data <- heatmap_data[, !colnames(heatmap_data) %in% 'Geneid', drop = TRUE]
                        #This won't work if heatmap_data is a tibble, must be a data.frame, so annoying!!

                        if(!is.null(gene_ordering_vector)){
                          #Remove genes not in the main data frame but in the ordering vector, 
                          # and then order based on the vector
                          remove_index <- rownames(heatmap_data) %in% gene_ordering_vector
                          heatmap_data <- heatmap_data[remove_index,,drop = FALSE]
                          heatmap_data <- heatmap_data[na.omit(
                            match(gene_ordering_vector, 
                                  rownames(heatmap_data))
                          ),,
                          drop=FALSE]
                        }

                        geneids <- rownames(heatmap_data)

                        #Get max/min TPM scores for the side annotation heatmap                       
                        tpm_mat <- list_element$tpm[match(rownames(heatmap_data),
                                                          list_element$tpm$Geneid),] %>% 
                          select(-Geneid) %>% 
                          as.matrix %>% 
                          log1p
                        rownames(tpm_mat) <- geneids

                        tpm_min <- rowMins(tpm_mat)
                        names(tpm_min) <- geneids
                        tpm_max <- rowMaxs(tpm_mat)
                        names(tpm_max) <- geneids
                        return(list(tpm_max = tpm_max,
                                    tpm_min = tpm_min))
                      })

    tpm_max_all <- mround(
      max(
        unlist(
          lapply(max_tpm, 
                 getElement,
                 'tpm_max')
        )
      ), 
      base = 1)

    tpm_breaks <- round(
      c(0,
        tpm_max_all/4,
        tpm_max_all/2,
        tpm_max_all*(3/4),
        tpm_max_all),
      digits = 2
    )
  }

  #Generate the colour function from the TPM legend breaks for the TPM annotation
  if(tpm_mat){
    tpm_col_fun <- colorRamp2(
      breaks = tpm_breaks,
      colors = viridis(length(tpm_breaks))
    )
  }

  if(is.numeric(threshold_tpm)){
    #Apply a threshold to the TPM in order to remove low expressed genes.
    # Remember to add log1p here! 
    # Threshold uses the max value, so it ony needs to be above thresh 
    #   in one stage avg in one transmission method.
    #Will alter the gene_filter vector in order to only select the genes above threshold

    gene_maxs <- bind_rows(
      lapply(max_tpm, 
             getElement,
             'tpm_max')
    )

    remove_genes <- names(gene_maxs)[apply(
      gene_maxs, 
      2, 
      function(maxs){
        all(maxs < log2(1+threshold_tpm))
      }
    )]

    filter_genes_vector <- filter_genes_vector[!filter_genes_vector %in% remove_genes]

  }

  if(is.character(annotation_clade)){
    order_cir <- gene_ordering_vector[gene_ordering_vector %in% filter_genes_vector]
    clade_info <- getElement(cir_info,
                             annotation_clade)[match(order_cir, 
                                                     cir_info$Geneid)]
    names(clade_info) <- str_split_fixed(order_cir, 
                                         pattern = '_', 
                                         n = 2)[,2]
    annotation_col <- annotation_col
  }

  heatmaps <- lapply(zscore_list, 
                     function(list_element){
                       #list_element <- zscore_list[[1]]
                       heatmap_data <- as.data.frame(list_element$zscores)
                       title <- list_element$experiment

                       if(!'Geneid' %in% colnames(heatmap_data)){
                         stop('No Geneid column in the heatmap data.frame')
                       }

                       if(!is.null(filter_genes_vector)){
                         heatmap_data <- as.data.frame(
                           filter(heatmap_data, 
                                  Geneid %in% filter_genes_vector)
                         )
                       }

                       rownames(heatmap_data) <- as.character(heatmap_data$Geneid)

                       heatmap_data <- heatmap_data[, !colnames(heatmap_data) %in% 'Geneid', drop = TRUE]
                       #This won't work if heatmap_data is a tibble, must be a data.frame, so annoying!!

                       if(!is.null(gene_ordering_vector)){
                         #Remove genes not in the main data frame but in the ordering vector,
                         #  and then order based on the vector
                         remove_index <- rownames(heatmap_data) %in% gene_ordering_vector
                         heatmap_data <- heatmap_data[remove_index,,drop = FALSE]
                         heatmap_data <- heatmap_data[na.omit(
                           match(gene_ordering_vector, 
                                 rownames(heatmap_data))
                         ),,
                         drop=FALSE]
                       }

                       if(tpm_mat){
                         #Get max/min TPM scores for the side annotation heatmap                       
                         tpm_mat <- list_element$tpm[match(rownames(heatmap_data),
                                                           list_element$tpm$Geneid),] %>%
                           select(-Geneid) %>% 
                           as.matrix %>% 
                           log1p

                         tpm_min <- rowMins(tpm_mat)
                         tpm_max <- rowMaxs(tpm_mat)

                         names(tpm_min) = rownames(heatmap_data)
                         names(tpm_max) = rownames(heatmap_data)

                         tpm_ha <- rowAnnotation(
                           min.max = anno_simple(cbind(tpm_min, 
                                                       tpm_max), 
                                                 col = tpm_col_fun,
                                                 gp = gpar(col = 'black',
                                                           lwd = 0.1),
                                                 border = TRUE,
                                                 simple_anno_size = unit(0.2, "cm")),
                           show_annotation_name = FALSE)
                       }

                       #Draw the left annotation is needed, only for the first element.

                       if(
                         identical(list_element,
                                   dplyr::first(zscore_list)) & is.character(annotation_clade)
                       ){
                         clade_ha <- rowAnnotation(
                           'subfamily' = anno_simple(
                             clade_info,
                             col = annotation_col
                           )
                         )
                       }

                       #Change rownames to remove PCHAS_
                       rownames(heatmap_data) = str_split_fixed(rownames(heatmap_data), 
                                                                pattern = '_', 
                                                                n = 2)[,2]

                       if(use_times){
                         colnames(heatmap_data) <- times_stages_table$time[match(colnames(heatmap_data),
                                                                                 times_stages_table$r_stage_diff_count)]
                       }

                       if(use_order_vector){
                         #Order the columns by ordering_vector
                         heatmap_data <- heatmap_data[,order(match(colnames(heatmap_data), 
                                                                   order_vector)),
                                                      drop = FALSE]
                       }else{
                         heatmap_data <- heatmap_data[,order(colnames(heatmap_data)),
                                                      drop = FALSE] 
                       }

                       h = Heatmap(heatmap_data,
                                   name = paste0('z-score',
                                                 extra_title),
                                   col = c(cbPalette[6],
                                           'white', 
                                           cbPalette[7]),
                                   row_names_gp = gpar(fontsize = ifelse(
                                     nrow(heatmap_data) > 13, 
                                     12*(13/nrow(heatmap_data)), 
                                     12)
                                   ),
                                   column_names_gp = gpar(fontsize = 10),
                                   row_dend_width = unit(3, "cm"),
                                   right_annotation = if(tpm_mat){
                                     tpm_ha
                                   }else{
                                     NULL
                                   },
                                   left_annotation = 
                                     if(identical(list_element,
                                                  dplyr::first(zscore_list)) & is.character(annotation_clade)){
                                       clade_ha
                                     }else{
                                       NULL
                                     }, 
                                   #Only want the clade annotation when it is first heatmap
                                   width = unit(4.5, 'cm'),
                                   cluster_columns = FALSE,
                                   height = unit(7,'cm'),
                                   row_title = NULL,
                                   cluster_rows = FALSE,
                                   column_title = paste(title),
                                   heatmap_legend_param = list(
                                     at = c(3,2,1,0,-1,-2,-3),
                                     labels_gp = gpar(fontsize = 10),
                                     title_gp = gpar(fontsize = 10)
                                   ),
                                   ...
                       )
                       return(h)
                     })

  ht_list <- Reduce('+', heatmaps) #To add the list elements together

  if(tpm_mat){
    #Draw the legend
    tpm_lgd <- Legend(
      title = "Min/Max log(TPM + 1)", 
      col = tpm_col_fun, 
      at = tpm_breaks, 
      labels = tpm_breaks,
      title_gp = gpar(fontsize = 7)
    )
  }

  if(is.character(annotation_clade)){
    clade_lgd <- Legend(
      title = 'sub-family',
      at = names(annotation_col),
      legend_gp = gpar(fill = annotation_col)
    )
  }

  draw(ht_list, 
       row_title_side = 'left',
       column_title_side = 'bottom',
       heatmap_legend_side = 'left',
       #Only include the annotation legends for the annotations we actually have.
       annotation_legend_list = if(is.character(annotation_clade) & tpm_mat){
         list(tpm_lgd,
              clade_lgd)
       }else if(tpm_mat & !is.character(annotation_clade) ){
         list(tpm_lgd)
       } else if(!tpm_mat & is.character(annotation_clade) ){
         list(clade_lgd)
       })
}
```

## Phaseograms of the entire genome.

```{r phaseograms_whole_genome}

draw_heatmap_zscore(zscore_list = twentyfour_z,
                    gene_ordering_vector = phaseogram_time_order$rMT$zscore_order$Geneid,
                    use_times = TRUE,
                    use_order_vector = FALSE,
                    show_row_names = FALSE,
                    tpm_mat = FALSE)

## Phaseograms of the _cir_ genes with clade annotations. - Supplementary figure 6

```{r pir_gene_phaseogram_breaks}
  max_tpm <- lapply(twentyfour_z, 
                    function(list_element){
                      #list_element <- zscore_list[[1]]
                      heatmap_data <- as.data.frame(list_element$zscores)
                      title <- list_element$experiment

                      if(!'Geneid' %in% colnames(heatmap_data)){
                        stop('No Geneid column in the heatmap data.frame')
                      }

                            heatmap_data <- as.data.frame(
                              filter(heatmap_data, 
                                     Geneid %in% cir_id)
                            )

                          rownames(heatmap_data) <- as.character(heatmap_data$Geneid)

                          heatmap_data <- heatmap_data[, !colnames(heatmap_data) %in% 'Geneid', drop = TRUE]
                          #This won't work if heatmap_data is a tibble, must be a data.frame, so annoying!!
                          # 
                          # if(!is.null(gene_ordering_vector)){
                          #   #Remove genes not in the main data frame but in the ordering vector, 
                          #   # and then order based on the vector
                          #   remove_index <- rownames(heatmap_data) %in% gene_ordering_vector
                          #   heatmap_data <- heatmap_data[remove_index,,drop = FALSE]
                          #   heatmap_data <- heatmap_data[na.omit(
                          #     match(gene_ordering_vector, 
                          #           rownames(heatmap_data))
                          #   ),,
                          #   drop=FALSE]
                          # }

                          geneids <- rownames(heatmap_data)

                          #Get max/min TPM scores for the side annotation heatmap                       
                          tpm_mat <- list_element$tpm[match(rownames(heatmap_data),
                                                            list_element$tpm$Geneid),] %>% 
                            select(-Geneid) %>% 
                            as.matrix %>% 
                            log1p
                          rownames(tpm_mat) <- geneids

                          tpm_min <- rowMins(tpm_mat)
                          names(tpm_min) <- geneids
                          tpm_max <- rowMaxs(tpm_mat)
                          names(tpm_max) <- geneids
                          return(list(tpm_max = tpm_max,
                                      tpm_min = tpm_min))
                        })

  tpm_max_all <- mround(
    max(
      unlist(
        lapply(max_tpm, 
               getElement,
               'tpm_max')
      )
    ), 
    base = 1)

  tpm_breaks <- round(
    c(0,
      tpm_max_all/4,
      tpm_max_all/2,
      tpm_max_all*(3/4),
      tpm_max_all),
    digits = 2
  )
```

```{r phaseograms_cir}

draw_heatmap_zscore(zscore_list = twentyfour_z,
                    gene_ordering_vector = phaseogram_time_order$rMT$zscore_order$Geneid,
                    filter_genes_vector = cir_id,
                    annotation_clade = 'SorL',
                    annotation_col = SorL_col,
                    threshold_tpm = 1,
                    use_times = TRUE,
                    use_order_vector = FALSE,
                    show_row_names = FALSE,
                    tpm_mat = FALSE)

draw_heatmap_zscore(zscore_list = twentyfour_z,
                    gene_ordering_vector = phaseogram_time_order$SBP$zscore_order$Geneid,
                    filter_genes_vector = cir_id,
                    annotation_clade = 'SorL',
                    annotation_col = SorL_col,
                    threshold_tpm = 1,
                    use_times = TRUE,
                    use_order_vector = FALSE,
                    show_row_names = FALSE,
                    tpm_breaks = tpm_breaks,
                    tpm_mat = FALSE)

## Figure 4 and supplementary figure 7/8

_pir_ genes selected by being expressed above 1 TPM in all samples.

Adding in housekeeping genes for comparison:
HSP70 putative  PCHAS_0721000
MSP1 PCHAS_0831300
AMA1 PCHAS_0931000

```{r pirs_side-by-side_figure4_RMT-SBP-completelySeparate}

housekeeping_genes <- data.frame('HSP70' = 'PCHAS_0721000',
                             'MSP1' = 'PCHAS_0831300',
                             'AMA1' = 'PCHAS_0931000')

cir_housekeeping_id <- c(cir_id, 
                         'PCHAS_0721000',
                         'PCHAS_0831300',
                         'PCHAS_0931000')

cir_plot <- twentyfour_ct_comb %>% 
  filter(Geneid %in% cir_housekeeping_id) %>%
  melt(id.vars = 'Geneid', 
       variable.name = 'sample', 
       value.name = 'tpm') %>% 
  mutate(
    stage = gsub(sample, pattern = '_[[:digit:]]$', replacement = ''),
    transmission = str_split_fixed(sample, pattern = '_', n = 3)[,1],
    diff_count = str_split_fixed(sample, pattern = '_', n = 4)[,3],
    time = str_split_fixed(sample, pattern = '_', n = 3)[,2],
    subfam = cir_clade_info$clade[match(Geneid, cir_clade_info$Geneid)]
  ) %>% 
  mutate(subfam = ifelse(is.na(subfam), 'ref', subfam)) %>% 
  filter(!(time %in% c('sch.culture', 
                       'Unknown'))) %>% 
  mutate(subfam = str_replace(subfam, 
                              'ancestral',
                              'a')) %>% 
  mutate(
    subfam = interaction(
      transmission,
      forcats::fct_shift(
        relevel(
          as.factor(subfam),
          ref = 'ref')
      ),
      sep = '_',
      lex.order = TRUE
    )
  )
#Need to add the transmission name to the subfam so that they are ordered correctly.
# Need to move 'ref' factor to the start of the subfam, then shift the factor levels to the 'left' (so now
# ref is at the end) ensuring that the 'ref' genes are at the end of the graph for each transmission type.

cir_plot$Geneid <- ifelse(
  cir_plot$Geneid %in% housekeeping_genes[1,],
  colnames(housekeeping_genes)[ match(cir_plot$Geneid, housekeeping_genes[1,]) ],
  cir_plot$Geneid
)

limits_cir <- c(0, 
                mround(max(cir_plot$tpm), 
                       base = 100))

cir_plot_order <- cir_plot %>% 
  filter(time == '14h',
         transmission == 'rMT') %>% 
  arrange(desc(tpm))

cir_plot <- mutate(cir_plot, 
                   Geneid = factor(Geneid, 
                                   levels = unique(cir_plot_order$Geneid)))

transmissionSep_cir_plot <- function(time_vec = '14h',
                                     log_it = TRUE,
                                     limits = limits_cir,
                                     point_size = 1,
                                     text_size = 10,
                                     threshold_min = 100,
                                     threshold_max = Inf){

  cir_plot_filter <- group_by(cir_plot, 
                       Geneid) %>% 
    filter(
      time == time_vec,
      !Geneid %in% cir_id | any(tpm >= threshold_min),
      !Geneid %in% cir_id | all(tpm <= threshold_max)
    )
  #Filter for pirs expressed above or below the thresholds, have the OR so that the ref genes aren't removed.
  # e.g. Either 'Not a cir' OR 'Any TPM is above the minimum'.

  # cir_plot_filter %>% group_by(Geneid) %>% summarise(n())
  if(log_it){
    cir_plot_filter$tpm <- log2(1+cir_plot_filter$tpm)
    limits = log2(1+limits)
    breaks <- round(seq.int(0, 
                            max(limits), 
                            length.out = round(max(limits)) + 1, 
                            digits = 0), 
                    digits = 0)
  }else{
    breaks <- seq.int(0, max(limits), length.out = 11)
  }

  cir_plot_filter$Geneid <- factor(
    str_remove(cir_plot_filter$Geneid, 
               pattern = 'PCHAS_'),
    levels = str_remove(levels(cir_plot_filter$Geneid), 
                        pattern = 'PCHAS_')
  )

  lapply(unique(cir_plot_filter$transmission), 
         function(transmission_method){

           cir_plot_filter <- filter(cir_plot_filter, 
                              transmission %in% transmission_method)

           y <- ggplot(
             data = cir_plot_filter,
             mapping = aes(
               x = Geneid,
               y = tpm
             )
           ) +
             stat_summary(fun.y = median, 
                          fun.ymin = median, 
                          fun.ymax = median,
                          geom = "bar",
                          fill = 'lightblue',
                          width = 0.75, 
                          size = 0.25, 
                          position = 'dodge') +
             geom_point(stat = 'identity', 
                        position = position_dodge(width = 0.75),
                        size = point_size) +
             scale_y_continuous(breaks = breaks,
                                limits = limits) +
             facet_grid(.~subfam, 
                        scales = "free_x",
                        space = "free_x",
                        labeller = labeller(subfam = function(string) {
                          str_remove_all(string, 
                                         pattern = '(rMT|SBP)_')
                        })) +
             theme_classic() +
             ylab('log(TPM + 1)')+
             ggtitle(paste(unique(cir_plot_filter$time)))

>>>>

---
title: "Manuscript_Github_Pchabaudi_calculateTPM"
author: "Timothy Little"
date: "22/03/2021"
output: html_document
---

@@

```{r read_tpm_rescaled_data}

chab_count_table <- read_xlsx('Manuscript_Frontiers_chabaudi_24h_counts.xlsx')

twentyfour_ct_comb <- tpm(chab_count_table, 
                          lengths = chabaudi_transcript_lengths)

#Setting twentyfour_tpm_ct core dataset - remove pseudogene and re-scaled

twentyfour_tpm_nopseudo <- twentyfour_ct_comb %>% 
  filter(!Geneid %in% 'PCHAS_0420800')

twentyfour_tpm_rescale_nopseudo <- lapply(colnames(twentyfour_tpm_nopseudo), 
                                          function(column_name){
                                            column <- getElement(twentyfour_tpm_nopseudo, 
                                                                 paste(column_name))
                                            if(is.numeric(column)){
                                              #is.numeric to avoid trying to scale the Geneid column
                                              tpm_scale_factor <- 1e6 / sum(column)
                                              tpm_scaled <- as.numeric(column*tpm_scale_factor)
                                            }else{
                                              tpm_scaled <- column
                                            }
                                            return(tpm_scaled)
                                          }) %>% 
  bind_cols

colnames(twentyfour_tpm_rescale_nopseudo) <- colnames(twentyfour_tpm_nopseudo)
# apply(twentyfour_tpm_rescale_nopseudo[,-1], 2, sum)

#Setting the re-scaled version as the default from now on
#Setting twentyfour_tpm_ct core dataset - removed pseudogene and re-scaled
twentyfour_ct_comb <- twentyfour_tpm_rescale_nopseudo

twentyfour_rMT <- select(twentyfour_ct_comb, 
                         matches('Geneid|rMT')) %>% 
  rename_with(~ str_extract(.x,pattern = '(?<=rMT_).+'),
              matches('rMT'))

twentyfour_rMT_avg <- melt(twentyfour_rMT, 
                           variable.name = 'sample',
                           value.name = 'tpm') %>% 
  mutate(stage = str_extract(sample, pattern = '.+(?=_[[:digit:]])')) %>% 
  group_by(Geneid, stage) %>% 
  summarise(mean_tpm = mean(tpm)) %>% 
  dcast(Geneid ~ stage)

# twentyfour_SBP <- select(twentyfour_ct_comb, 
#                          matches('Geneid|SBP')) %>% 
#   rename_with(~ str_extract(.x,pattern = '(?<=SBP_).+'),
#               matches('SBP'))
# 
# twentyfour_SBP_avg <- melt(twentyfour_SBP, 
#                            variable.name = 'sample',
#                            value.name = 'tpm') %>% 
#   mutate(stage = str_extract(sample, pattern = '.+(?=_[[:digit:]])')) %>% 
#   group_by(Geneid, stage) %>% 
#   summarise(mean_tpm = mean(tpm)) %>% 
#   dcast(Geneid ~ stage)

twentyfour_avgtpm <- list(
  'rMT' = list(
    tpm = twentyfour_rMT_avg,
    experiment = 'rMT'
  # ),
  # 'SBP' = list(
  #   tpm = twentyfour_SBP_avg,
  #   experiment = 'SBP'
  )
)

@@
